\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-23 k-planar.tex
Big reorganization done

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a 'universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $p$-centered colouring \cite{DFMS21}, $\ell$-vertex ranking~\citep{BDJM}, twin-width~\citep{BKW}, and odd colourings~\citep{DMO}.

@@

All of these results hold for any graph class that has a product structure theorem analogous to \cref{PlanarProduct}; that is, for any graph class  $\mathcal{G}$ where every graph in $\mathcal{G}$ is contained in $H\boxtimes P\boxtimes K_\ell$ where $H$ has bounded treewidth, $P$ is a path, and $\ell$ is bounded.\footnote{It is easily seen that $\tw(H\boxtimes K_\ell) \leq (\tw(H)+1)\ell-1$, so we may assume that $\ell=1$ in this statement.} These applications motivate finding product structure theorems for other graph classes. \citet{DJMMUW20} prove product structure theorems for graphs of bounded Euler genus\footnote{The \textit{Euler genus} of a surface with $h$ handles and $c$ crosscaps is $2h+c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} and for apex-minor-free graphs,\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \defin{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \defin{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \defin{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.} and \citet{DEMWW22} do so for graphs in any minor-closed class and with bounded maximum degree. See \cref{Generalisations} for more precise statements and see \citep{DHJLW21} for a survey on this topic.

@@

\subsection{$k$-Planar Graphs}
\label{k_planar_section}

We start with the example of $k$-planar graphs. A graph is \defin{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings, where no three edges cross at a single point (see \cref{sec-k-planar} for a formal definition). Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. It is well-known that the family of $k$-planar graphs is not minor-closed.  Indeed, 1-planar graphs may contain arbitrarily large complete graph minors~\citep{dujmovic.eppstein.ea:structure}. Hence the above results are not applicable for  $k$-planar graphs. We extend \cref{PlanarProduct} as follows.

@@

For the important special case of $1$-planar graphs, we are able to obtain better constants and an additional property (planarity) on $H$.

@@

For at least one application, the planarity of $H$ is critical in obtaining asymptotically tight bounds \cite{BDJM}. \cref{1_planar_product} is proven in the more general setting of $d$-framed multigraphs. This more general result implies a product structure theorem for $d$-map graphs in which $H$ is also planar.

@@

Although $k$-planar graphs are the most high-profile target for a generalization of \cref{PlanarProduct}, we actually prove a substantially stronger result than \cref{kPlanarProduct} using the following definition. A non-empty set $\SS$ of non-trivial paths in a graph $G$ is a \defin{$(k,d)$-shortcut system} (for $G$) if:

@@

The following theorem is one of the main contributions of the paper. It says that if a graph class $\mathcal{G}$ has a product structure theorem, then the class of graphs obtained by applying a shortcut system to graphs in $\mathcal{G}$ also has a product structure theorem.

@@

Theorem~\ref{PlanarProduct}(b), \cref{ShortcutProduct}, and \cref{AddDummy} imply \cref{kPlanarProduct} with $K_{6(k^3+3k)}$ instead of $K_{18k^2+48k+30}$. Some further observations presented in \cref{sec-k-planar} lead to the improved result.  \cref{ShortcutProduct} is applicable for many graph classes in addition to $k$-planar graphs. Some examples are explored in \cref{first_example,mid_example,last_example}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Layered Partitions}
\label{LayeredPartitions}

\note{PM}{Move this section out of the introduction, into a ``preliminaries'' section?}

While strong products enable concise statements of the theorems in \cref{Introduction}, to prove such results it is helpful to work with layerings and partitions, which we now introduce.

@@

As an example of the use of layered partitions, to prove \cref{PlanarProduct}(a),
\citet{DJMMUW20} showed that every planar graph has an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$ (improved to $6$ in \cite{UWY}). The proof is constructive and gives a simple quadratic-time algorithm for finding the corresponding partition and layering.\footnote{\citet{bose.morin.ea:optimal} have recently given linear time algorithms for computing layered $H$-partitions of planar graphs.} At the core of their work is the elegant proof by \citet{PS21} of the following result:

\begin{thm}[\citep{PS21}]
\label{ps}
  Every planar triangulation $G$ has an $H$-partition $\PP$ such that $\tw(H)\leq 8$ and $G[P]$ is a shortest path in $G$ for each $P\in\PP$.
\end{thm}

Indeed, the above-mentioned result of \citet{DJMMUW20} is a slight strengthening of \cref{ps}, where for each $P\in\PP$ no two vertices of $P$ have the same distance to some fixed root vertex $r$.

The following result is the main technical contribution of the paper.   Loosely speaking, it shows that if a graph $G$ has product structure, then so too does $G^\SS$ for every shortcut system $\SS$ of $G$.

\begin{restatable}{thm}{mmg}
	\label{ShortcutPartition}
	Let $G$ be a graph having an $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\SS$ be a $(k,d)$-shortcut system for $G$.  Then $G^\SS$ has a $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$.
\end{restatable}

Note that \cref{ShortcutPartition} is equivalent to \cref{ShortcutProduct} by \cref{PartitionProduct}.

\begin{table}
  \begin{center}
    \begin{tabular}{lccl}
      \textbf{Graph class} & \boldmath $\tw(H)$ & \boldmath $\ell$ & \textbf{Ref.} \\ \hline\hline
      Planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      Planar & $6^*$ & $1$ & \cite{UWY} \\
      Genus-$g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW} \\
      Genus-$g$ & $2g+6$ & 1 & \cite{UWY} \\
      Apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\[1ex]
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringPartition} \\
      $k$-NN & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $\textcolor{red}{?}$ & $\textcolor{red}{?}$ & \textcolor{red}{\cref{dMapProduct}} \\ \hline
      \\[.2ex]
      \multicolumn{4}{l}{$^*$ the bound also applies to the simple treewidth of $H$.}
    \end{tabular}
  \end{center}
  \caption{Summary of known product structure theorems.}
  \label{summary_table}
\end{table}

\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{Structure} to derive product structure theorems for string graphs, powers of bounded-degree graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive product structure theorems for $1$-planar graphs and map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure and the consequences of the current work.
  \item \cref{WrappingUp} concludes with an open problem.
\end{compactitem}

The purpose of this section is to prove \cref{ShortcutPartition} and its specialization to $k$-planar graphs \cref{kPlanarProduct}.

@@

\cref{ShortcutPartition} shows how, given a $(k,d)$-shortcut system $\SS$ of a graph $G$, an $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\SS}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.  For convenience, it will be helpful to assume that $\SS$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\SS$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\SS}$ but eliminates special cases in some of our proofs.

Consider a tree $T$ rooted at some node $x_0\in V(T)$. Each subtree $T'$ of $T$ is considered to be rooted at the node in $T'$ closest to $x_0$. A node $a\in V(T)$ is a \defin{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \defin{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \defin{strict} $T$-ancestor of $x$ if $a\neq x$.
The \defin{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  The \defin{lowest common ancestor} of a non-empty set $S\subseteq V(T)$ is the maximum $T$-depth node of $T$ that is a $T$-ancestor of every node in $S$.
% For each node $x\in V(T)$, define
% \[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
% to be the maximal subtree of $T$ rooted at $x$.

@@

\subsection{\boldmath $k$-Planar Graphs}

@@

\subsection{\boldmath $(g,k)$-Planar Graphs}

@@

\begin{thm}[\citep{DJMMUW20,UWY,DHHW}]\label{GenusProduct}
  Every graph of Euler genus $g$ is contained in:
  \begin{compactenum}[(a)]
  \item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
  \item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
  \end{compactenum}
\end{thm}

@@

In this section we give three examples of well-known graph families that can be expressed in terms of shortcut systems and which therefore have product structure theorems.

@@

A similar proof to that of \cref{StringShortcut} shows that every $(g,\delta)$-string graph is $(g,2\delta^2)$-planar. Theorems~\ref{PlanarProduct}(b), \ref{ShortcutProduct}  and \ref{GenusProduct}(b):
 % and \cref{StringShortcut,qn} imply:

\begin{thm}\label{StringPartition}
For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$ and $t:= \binom{ \delta+4}{3}-1$.
Then  every $(g,\delta)$-string graph is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
\end{thm}

@@

\begin{thm}
\label{gdMapProduct}
For integers $g\geq 0$ and $d\geq 3$, let $\ell:= 14\max\{2g,3\}\,\binom{d}{2}$.
Then every $(g,d)$-string graph is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $9$,
\end{thm}

@@

\begin{lem}
\label{MewMapGraphLemma}
For all integers $g\geq 0$ and $d\geq 3$, every $d$-map graph is a spanning subgraph of a $d$-framed graph.\todo{All I did here was remove references to $g$}
\end{lem}

@@

For each face $F$ of $G_1$ that corresponds to a vertex $x$ of $G_0$, the size of $F$ equals the number of nations incident to $x$, which is at most $d$.
So every face of $G_1$ has size at most $d$.  Let $\hat{G}_1$ be the $d$-framed graph whose frame is $G_1$.

To complete the proof, we now show that $\hat{G}_1$ contains $G$.
If $vw\in E(G)$ then the nation faces of $G_0$ corresponding to $v$ and $w$ have a vertex $x$ of $G_0$ in common.   If some face of $G_1$ corresponds to $x$ then $v$ and $w$ are on a common face of $G_1$ so $vw\in E(\hat{G}_1)$.  If no face of $G_1$ corresponds to $x$ then the only two nation faces of $G_0$ with $x$ on their boundary those corresponding to $v$ and $w$, in which case $vw\in E(G_1)\subseteq E(\hat{G}_1)$.  Therefore $G\subseteq\hat{G}_1$.
\end{proof}

@@

Combining \cref{non-repetitive} with our results on product structure, we obtain the following corollary:
\begin{cor}\label{nonrepetitive_cor}
  The following bounds hold for the nonrepetitive chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30)\cdot 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10) 4^{\binom{k+4}{3}}$
    % \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $1792$
    \item $d$-map: $256(d+3\floor{d/2}-3)$
    % \item $(g,d)$-map: $g4^{O(d)}$
    % \item $k$-NN: $4^{O(k^6)}$
    \item $(g,d)$-map graph: \textcolor{red}{??}
    % \item
  \end{compactenum}
\end{cor}

@@

\citet{PS21} and \citet{DFMS21} use \cref{ps} and \cref{PlanarProduct}(b), respectively, to show that $\chi_p(G)$ is polynomial in $p$ when $G$ is planar or of bounded Euler genus.
Upper bounds on $\chi_p$ for graphs of given treewidth~\citep{PS21} and for graph products~\citep{DFMS21} imply the next lemma.

>>>>

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to prove that planar graphs have $p$-centered chromatic number $O(p^2\log p)$ and give a matching lower bound.
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a 'universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $\ell$-vertex ranking~\citep{BDJM}, twin-width~\citep{BKW}, and odd colourings~\citep{DMO}.

@@

All of these results hold for any graph class that has a product structure theorem analogous to \cref{PlanarProduct}; that is, for any graph class  $\mathcal{G}$ where every graph in $\mathcal{G}$ is contained in $H\boxtimes P\boxtimes K_\ell$ where $H$ has bounded treewidth, $P$ is a path, and $\ell$ is bounded.\footnote{It is easily seen that $\tw(H\boxtimes K_\ell) \leq (\tw(H)+1)\ell-1$, so we may assume that $\ell=1$ in this statement.} These applications motivate finding product structure theorems for other graph classes. \citet{DJMMUW20} prove product structure theorems for graphs of bounded Euler genus\footnote{The \textit{Euler genus} of a surface with $h$ handles and $c$ crosscaps is $2h+c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} and for apex-minor-free graphs,\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \defin{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \defin{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \defin{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.} and \citet{DEMWW22} do so for bounded-degree graphs in any minor-closed class. See 
% \cref{Generalisations} for more precise statements and see 
\citep{DHJLW21} for a survey on this topic.

@@

\subsection{$k$-Planar Graphs}
% \label{k_planar_section}

We start with the example of $k$-planar graphs. A graph is \defin{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings, where no three edges cross at a single point (see \cref{k_planar_section} for a formal definition). Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. It is well-known that the family of $k$-planar graphs is not minor-closed.  Indeed, 1-planar graphs may contain arbitrarily large complete graph minors~\citep{dujmovic.eppstein.ea:structure}. Hence the above results are not applicable for  $k$-planar graphs. We extend \cref{PlanarProduct} as follows.

@@

For the important special case of $1$-planar graphs, we are able to obtain better constants and an additional property (planarity) of $H$.

@@

For at least two applications, the planarity of $H$ is critical in obtaining asymptotically tight bounds \cite{BDJM,DFMS21}. \cref{1_planar_product} is proven in the more general setting of $d$-framed multigraphs. This more general result implies a product structure theorem for $d$-map graphs in which $H$ is also planar.

@@

Although $k$-planar graphs are the most high-profile target for a generalization of \cref{PlanarProduct}, we actually prove a more general result than \cref{kPlanarProduct} using the following definition. A non-empty set $\SS$ of non-trivial paths in a graph $G$ is a \defin{$(k,d)$-shortcut system} (for $G$) if:

@@

The following theorem is one of the main contributions of the paper. It says that if a graph class $\mathcal{G}$ has a product structure theorem like \cref{PlanarProduct}(b), then so too does the class of graphs obtained by applying shortcut systems to graphs in $\mathcal{G}$.

@@

Theorem~\ref{PlanarProduct}(b), \cref{ShortcutProduct}, and \cref{AddDummy} imply \cref{kPlanarProduct} with $K_{6(k^3+3k)}$ instead of $K_{18k^2+48k+30}$. Some further observations presented in \cref{k_planar_section} lead to the improved result.  \cref{ShortcutProduct} is applicable for many graph classes in addition to $k$-planar graphs. Some examples are explored in \cref{first_example,mid_example,last_example}.

@@

\begin{table}
  \begin{center}
    \begin{tabular}{lccl}
      \textbf{Graph class} & \boldmath $\tw(H)$ & \boldmath $\ell$ & \textbf{Ref.} \\ \hline\hline
      Planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      Planar & $6^*$ & $1$ & \cite{UWY} \\
      Genus-$g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW} \\
      Genus-$g$ & $2g+6$ & 1 & \cite{UWY} \\
      Apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\[1ex]
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-NN & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \\[.2ex]
      \multicolumn{4}{l}{$^*$ the bound also applies to the simple treewidth of $H$.}
    \end{tabular}
  \end{center}
  \caption{Known product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for string graphs, powers of bounded-degree graphs, map graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure and the consequences of the current work.
  \item \cref{WrappingUp} concludes with an open problem.
\end{compactitem}

The purpose of this section is to prove \cref{ShortcutProduct} and its specialization to $k$-planar graphs, \cref{kPlanarProduct}.
While strong products enable concise statements of the theorems in \cref{Introduction}, to prove such results it is helpful to work with layerings and partitions, which we now introduce.

\subsection{Layered Partitions}

@@

As an example of the use of layered partitions, to prove \cref{PlanarProduct}(a),
\citet{DJMMUW20} build on an earlier result of \citet{PS21} to show that every planar graph has an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$ (improved to $6$ in \cite{UWY}). The proof is constructive and gives a simple quadratic-time algorithm for finding the corresponding partition and layering.\footnote{\citet{bose.morin.ea:optimal} have recently given linear time algorithms for computing layered $H$-partitions of planar graphs.}

% \note{PM}{This feels out of place now that this section has been moved out of the introduction.  Delete?
% 
% At the core of their work is the elegant proof by \citet{PS21} of the following result:
% 
% \begin{thm}[\citep{PS21}]
% \label{ps}
%   Every planar triangulation $G$ has an $H$-partition $\PP$ such that $\tw(H)\leq 8$ and $G[P]$ is a shortest path in $G$ for each $P\in\PP$.
% \end{thm}
% 
% Indeed, the above-mentioned result of \citet{DJMMUW20} is a slight strengthening of \cref{ps}, where for each $P\in\PP$ no two vertices of $P$ have the same distance to some fixed root vertex $r$.
% }

By \cref{PartitionProduct}, \cref{ShortcutProduct} is equivalent to the following result, whose proof is the subject of the next section.

\begin{thm}
	\label{ShortcutPartition}
	Let $G$ be a graph having an $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\SS$ be a $(k,d)$-shortcut system for $G$.  Then $G^\SS$ has a $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$.
\end{thm}

@@

We now prove \cref{ShortcutPartition}.
For convenience, it will be helpful to assume that $\SS$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\SS$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\SS}$ but eliminates special cases in some of our proofs.

Let $T$ be a tree rooted at some node $x_0\in V(T)$.  A node $a\in V(T)$ is a \defin{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \defin{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \defin{strict} $T$-ancestor of $x$ if $a\neq x$.
The \defin{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  The \defin{lowest common $T$-ancestor} of a non-empty set $S\subseteq V(T)$ is the maximum $T$-depth node of $T$ that is a $T$-ancestor of every node in $S$.  In the following, we treat any subtree $T'$ of $T$ as a rooted tree whose root is the lowest common $T$-ancestor of $V(T')$.

% For each node $x\in V(T)$, define
% \[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
% to be the maximal subtree of $T$ rooted at $x$.

@@

\subsection{\boldmath $k$-Planar Graphs}
\label{k_planar_section}

@@

\subsection{\boldmath $(g,k)$-Planar Graphs}
\label{gk_planar_section}

@@

\begin{thm}[\citep{DJMMUW20,UWY,DHHW}]\label{GenusProduct}
  Every graph of Euler genus $g$ is contained in:
  \begin{compactenum}[(a)]
  \item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
  \item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some planar graph $H$ of treewidth at most $3$ and for some path $P$.
  \end{compactenum}
\end{thm}

@@

In this section we give four examples of well-known graph families that can be expressed in terms of shortcut systems and which therefore have product structure theorems.

@@

% A similar proof to that of \cref{StringShortcut} shows that every $(g,\delta)$-string graph is $(g,2\delta^2)$-planar.

\cref{StringShortcut,ShortcutProduct,GenusProduct} now imply:
 % and \cref{StringShortcut,qn} imply:

\begin{thm}\label{StringProduct}
For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$ and $t:= \binom{ \delta+4}{3}-1$.
Then  every $(g,\delta)$-string graph is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
\end{thm}

@@

\begin{thm}
\label{gdMapProduct}
For integers $g\geq 0$ and $d\geq 3$, let $\ell:= \max\{2g,3\}(7d^2-21d)$.
Then every $(g,d)$-map graph is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $9$,
\end{thm}

@@

\note{PM}{Got rid of genus-$g$ in the following. We use shortcuts for $(g,d)$-map graphs and now we don't have to define $(g,d)$-frame.}

\begin{lem}
\label{MewMapGraphLemma}
For all integers $g\geq 0$ and $d\geq 3$, every $d$-map graph is a spanning subgraph of a $d$-framed graph.
\end{lem}

@@

For each face $F$ of $G_1$ that corresponds to a vertex $x$ of $G_0$, the size of $F$ equals the number of nations incident to $x$, which is at most $d$.
So every face of $G_1$ has size at most $d$.  Let $\widehat{G}_1$ be the $d$-framed graph whose frame is $G_1$.

To complete the proof, we now show that $\widehat{G}_1$ contains every edge of $G$.
If $vw\in E(G)$ then the nation faces of $G_0$ corresponding to $v$ and $w$ have a vertex $x$ of $G_0$ in common.   If some face of $G_1$ corresponds to $x$ then $v$ and $w$ are on a common face of $G_1$ so $vw\in E(\widehat{G}_1)$.  If no face of $G_1$ corresponds to $x$\footnote{This can occur when $x$ is on the boundary of exactly two nation faces and two lake faces.} then the only two nation faces of $G_0$ with $x$ on their boundary are those corresponding to $v$ and $w$, in which case $vw\in E(G_1)\subseteq E(\widehat{G}_1)$.  Therefore $G\subseteq\widehat{G}_1$.
\end{proof}

@@

Combining \cref{non-repetitive} with our results on product structure, we obtain the following corollary:
\begin{cor}\label{nonrepetitive_cor}
  The following bounds hold for the nonrepetitive chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30)\cdot 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10) 4^{\binom{k+4}{3}}$
    % \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $1792$
    \item $d$-map: $256(d+3\floor{d/2}-3)$
    % \item $(g,d)$-map: $g4^{O(d)}$
    % \item $k$-NN: $4^{O(k^6)}$
    \item $(g,d)$-map graph: $(7d^2-21d)\cdot 4^{10}$
    % \item
  \end{compactenum}
\end{cor}

@@

% \citet{PS21} and 
\citet{DFMS21} use \cref{PlanarProduct} and \cref{PlanarProduct}(b), respectively, to show that $\chi_p(G)\in O(p^2\log p)$ when $G$ is planar or of bounded Euler genus.  Upper bounds on $\chi_p$ for graphs of given treewidth~\citep{PS21} and for graph products~\citep{DFMS21} imply the next lemma.
